---
title: "simple example for conditioanl probability"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## data summary

We have a hypothetical population of 1 million individuals with the following conditional probabilities as described below:

The test is positive 85% of the time when tested on a patient with the disease (high sensitivity):  ğ‘ƒ(test+|disease)=0.85 
The test is negative 90% of the time when tested on a healthy patient (high specificity):  ğ‘ƒ(testâˆ’|heathy)=0.90 
The disease is prevalent in about 2% of the community:  ğ‘ƒ(disease)=0.02



```{r sample dataset for conditional probability}
set.seed(1, sample.kind="Rounding") 

disease <- sample(c(0,1), size=1e6, replace=TRUE, prob=c(0.98,0.02))
test <- rep(NA, 1e6)
test[disease==0] <- sample(c(0,1), size=sum(disease==0), replace=TRUE, prob=c(0.90,0.10))
test[disease==1] <- sample(c(0,1), size=sum(disease==1), replace=TRUE, prob=c(0.15, 0.85))

head(test)

```
```{r summary of conditional probability calculations}
p_of_having_disease<-mean(disease==1)
p_of_not_having_disease<-mean(disease==0)
p_of_negative_test<-mean(test==0)
p_of_positive_test<-mean(test==1)
p_negtest_given_decease<-mean(test[disease==1]==0)
p_postest_given_decease<-mean(test[disease==1]==1)
p_having_disease_given_neg_test<-p_negtest_given_decease * (p_of_having_disease/p_of_negative_test)
p_having_disease_given_pos_test<-p_postest_given_decease * (p_of_having_disease/p_of_positive_test)
#If a patient's test is positive, how much does that increase their risk of having the disease?
p_of_having_disease
p_of_not_having_disease
p_of_negative_test
p_of_positive_test
p_having_disease_given_neg_test
p_having_disease_given_pos_test
#If a patient's test is positive, how much does that increase their risk of having the disease?
p_having_disease_given_pos_test/p_of_having_disease




```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
